Certified Knowledge Compilation \\ with Application to Verified Model Counting

### 2023-03-13 doc/lean.tex
start paper section on the verified checker

>>>>

[NOTE (JA): I am not worrying about length for the moment. This entire section can become an appendix, and then we can give an executive summary using whatever space
we have available]

We have implemented the rightmost two components of Figure~\ref{fig:chain}, namely the
proof checker and both model and weighted model counters, in the Lean 4 programming language
and proof assistant [cite]. Lean 4 implements a logical foundation in which expressions
have a computational interpretation, and just as with other proof assistants like Isabelle and Coq,
a function defined within the formal system can be compiled to efficient code.
At the same time, we can state and prove claims about the function within the system, thereby
verifying that the functions compute the intended results.

Verifying our checker and the model counters in this way serves at least two purposes.
When we run them on particular data sets, they provide high confidence
that the particular results we obtain are correct.
More broadly, however, our success also validates our general approach,
showing, in particular, that the CRAT proof format is sound.
Indeed, our formalization helped us discover a bug in our implementation
of the informal proof checker: in the ``\emph{s}'' step described at the end of Subsection~\ref{subsection:syntax}, it is sound (and necessary) to allow defining
clauses for the new operations in the proof of $\obar{\lit}_1 \lor \obar{\lit}_2$,
but it is generally not sound to allow the input clauses or other added clauses. [NOTE: let's add this specification to the end of Section 6.1.]
Our verified implementation of the checker tells us that the algorithm is sound and provides a reference for other implementations.

In this section, we describe the implementation in Lean and the specifications we have proved.
Lean 4's logical foundation describes a functional programming language with
inductive data types,
but our procedures also make use of calculations with natural numbers, integers, and rationals,
and data structures for arrays and hashmaps.
Lean's code extraction uses efficient but unverified versions of these,
and so in Subsection~\ref{subsection:verification:specifics} we clarify
what exactly we have verified and what we currently trust.

\subsection{Logical objects and operations}

Our specifications are built on a generic library for the syntax and semantics of
propositional logic. The data type of propositional formulas with variables indexed by
a datatype \lstinline{ν} is defined inductively as follows:
\begin{lstlisting}
inductive PropForm (ν : Type)
  | var (x : ν)
  | tr
  | fls
  | neg (φ : PropForm ν)
  | conj (φ₁ φ₂ : PropForm ν)
  | disj (φ₁ φ₂ : PropForm ν)
  | impl (φ₁ φ₂ : PropForm ν)
  | biImpl (φ₁ φ₂ : PropForm ν)
\end{lstlisting}
The last two constructors denote implication and bi-implication, which are not used in this
project. The usual syntactic and semantic notions are defined in the expected ways.
It is convenient to take a truth assignment \lstinline{PropAssignment ν} to a function that
assigns a boolean value to every assignment, and later restrict attention to subsets of the
variables. In Lean, we introduce the notation \lstinline{τ ⊨ φ} for the statement that
\lstinline{τ} satisfies \lstinline{φ}. We write \lstinline{entails φ₁ φ₂} to say that
\lstinline{φ₁} and \lstinline{φ₂} and \lstinline{equivalent φ₁ φ₂}.
For the purpose of model counting, if \lstinline{s} is any finite set of variables,
we define \lstinline{models φ s} to be the finite set of truth assignments \lstinline{τ}
satisfying \lstinline{φ} that assign arbitrary values on \lstinline{s} and
\lstinline{false} outside \lstinline{s}.

In logic, the set of propositional formulas modulo equivalence is sometimes called the \emph{Lindenbaum--Tarski algebra}. We have found it convenient to define this quotient,
which we denote \lstinline{PropTerm ν}, and lift the boolean operations and the entailment
relation to this new type. The advantage is that equivalent formulas give
rise to equal elements in the quotient.
This makes easier to prove equivalences as we go, since we can calculate with the associated
elements of the quotient and substitute one for another when the corresponding formulas are equivalent.

In our application, we instantiate \lstinline{ν} to a type \lstinline{Var} consisting
of positive natural numbers, corresponding to the indexing of variables in the DIMACS
format for SAT. It may be helpful to think of the elements of \lstinline{PropForm Var},
\lstinline{PropTerm Var}, and \lstinline{PropAssignment Var} as mathematical objects
rather than computational data. Our proof checker and model counters do not compute
with them; rather, we use them to specify what the checker and model counters are supposed to do.
The starting point for the proof checker is, instead, a CNF formula, which \emph{is}
presented as data.
Specifically, a literal is a nonzero integer, with the understanding that $-\ell$ represents
the negation of $\ell$. A clause is an array of literals, and a CNF formula is an array
of clauses:
\begin{lstlisting}
def ILit := { i : Int // i ≠ 0 }
abbrev IClause := Array ILit
abbrev ICnf := Array IClause
\end{lstlisting}
We define operations that translate these objects to the elements of \lstinline{PropForm Var}
that they represent. Lean provides us with helpful ``anonymous projection'' syntax that allows
us to write, for example, \lstinline{C.toPropForm} for the expression
\lstinline{Clause.toPropForm C}.
Similarly, we have a representation \lstinline{C.toPropTerm}
of \lstinline{C} as an equivalence class of propositional formulas, and we can speak of the
value of \lstinline{C} under any truth assignment.
Our code represents partial truth assignments as hashmaps:
\begin{lstlisting}
abbrev PartPropAssignment := HashMap Var Bool
\end{lstlisting}
We have written a procedure that reduces a clause \lstinline{C} with respect to a partial
assignment, and another procedure that uses a hashset to test whether a clause \lstinline{C} is a
tautology.
We have also verified that these low-level procedures do what they are supposed to do.

\subsection{Proof Checker}

Our proof checker interprets the input clauses listed in a CRAT file as a CNF formula $\varphi$.
It then processes and checks each rule, throwing an exception if a precondition is not met.
If it terminates without raising an exception, it outputs a POG, $P$, with distinguished root
$\bf r$.
We have formally proved the claim that in this case the formula corresponding to node $\bf r$ is a partitioned formula and is
logically equivalent to $\varphi$.
The POG $P$ can be saved to a file [TODO: implement this?] or passed on to one of the model counting
procedures described in the next section.

In our code, each element of a POG  is either a variable, a binary disjunction, or an arbitrary conjunction:
\begin{lstlisting}
inductive PogElt where
  | var  : Var → PogElt
  | disj : Var → ILit → ILit → PogElt
  | conj : Var → Array ILit → PogElt
\end{lstlisting}
In the first case, the argument \lstinline{x} in the expression
\lstinline{var x} is the index
of the variable; in \lstinline{disj x left right} and \lstinline{conj x args}
it is the definition number in the CRAT file.
The CRAT generator described in Section~\ref{section:generating:crat}
declares new variables sequentially, so we currently make the simplifying assumption
that the $i$th element of the array has the node with index $i + 1$,
but it will not be hard to generalize this.
Notice that the arguments to \lstinline{disj} and \lstinline{conj} are literals,
corresponding to positive or negative instances of variables declared earlier in the POG.
So our POG data structure is as follows:
\begin{lstlisting}
structure Pog where
  elts : Array PogElt
  wf : ∀ i : Fin elts.size, elts[i].args_decreasing
  inv : ∀ i : Fin elts.size, i = elts[i].varNum.natPred
\end{lstlisting}
The invariant \lstinline{wf} says that the graph is well founded, in the
sense that if \lstinline{elts[i]} is a disjunction
or a conjunction, then the variables occurring in the arguments have smaller indices.
Note that \lstinline{Fin elts.size} is the data type consisting of natural numbers
less than the size of the array \lstinline{elts}; our code enforces statically
that the array accesses are within bounds.

Note that we have not yet said that the corresponding formula is partitioned.
For each POG \lstinline{P} and literal \lstinline{l}, we define
\lstinline{P.toPropForm l} to be the
propositional formula that arises from interpreting \lstinline{l} as a propositional
formula, unfolding all the defined conjunctions and disjunctions. We define
what it means for such a formula to be partitioned:
\begin{lstlisting}
def partitioned : PropForm ν → Prop
  | tr         => True
  | fls        => True
  | var _      => True
  | neg φ      => φ.partitioned
  | disj φ ψ   => φ.partitioned ∧ ψ.partitioned ∧
                    ∀ τ, ¬ (φ.eval τ ∧ ψ.eval τ)
  | conj φ ψ   => φ.partitioned ∧ ψ.partitioned ∧ (φ.vars ∩ ψ.vars = ∅)
  | impl _ _   => False
  | biImpl _ _ => False
\end{lstlisting}
In other words, we rule out formulas with implication and bi-implication, and otherwise we follow the straightforward recursive definition.

With these definitions in place, we can explain the specification for our checker.
Our parser reads a CRAT file and returns an input formula \lstinline{cnf : ICnf} and a CRAT proof {pf : Array CratStep}, that is, an array of single proof steps:
\begin{lstlisting}
inductive CratStep (α ν β : Type)
  | /-- Add asymmetric tautology. -/
    addAt (idx : α) (C : Array β) (upHints : Array α)
  | /-- Delete asymmetric tautology. -/
    delAt (idx : α) (upHints : Array α)
  | /-- Declare product operation. -/
    prod (idx : α) (x : ν) (ls : Array β)
  | /-- Declare sum operation. -/
    sum (idx : α) (x : ν) (l₁ l₂ : β) (upHints : Array α)
  | /-- Delete operation. -/
    delOp (x : ν)
\end{lstlisting}
The core checker takes those as input, as well as a specification of the final node \lstinline{r}, and either throws an exception, indicating that the proof is not well formed, or returns a POG \lstinline{P}.
The specification asserts that, in the latter case, \lstinline{P.toPropForm r} is partitioned
and is equivalent to \lstinline{cnf.toPropForm}.

[TODO(JA): I'll leave stop here now because I am running out of steam today,
and also because Wojciech is still working on it.
But here is where we can sketch what the checker does.
For each line, it does some checks, it adds clauses to a clause database,
and it add a definition to the POG.
There is a list of invariants having to do with the original formula,
the clause database, and the POG, and we prove that if the checks are satisfied,
then at each step the invariants hold.
We should list the invariants and say something about how the checks are
performed.]

\subsection{The Model Counter}

The number of models that a propositional formula has depends on what we take the
variables to be. For example, there is only one assignment to the variables
$\{ x, y \}$ that makes $x \wedge y$ true, but there are two assignments to the
variables $\{ x, y, z \}$ that make $x \wedge y$ true.
It is not hard to show that for partitioned formulas \lstinline{PropForm ν} whose
variables are among a finite set \lstinline{s} of variables of cardinality \lstinline{numVars},
we can count the number of models of the formula on the variables in \lstinline{s}
recursively as follows:
\begin{lstlisting}
def countModels (nVars : Nat) : PropForm ν → Nat
  | tr         => 2^nVars
  | fls        => 0
  | var _      => 2^(nVars - 1)
  | neg φ      => 2^nVars - φ.countModels nVars
  | disj φ ψ   => φ.countModels nVars + ψ.countModels nVars
  | conj φ ψ   => φ.countModels nVars * ψ.countModels nVars / 2^nVars
  | impl _ _   => 0
  | biImpl _ _ => 0
\end{lstlisting}
Note that we can carry out this calculation for any formula at all, though the
calculation is meaningless if the formula is not partitioned.
Also recall that if the formula is partitioned, it does not contain implication or bi-implication.

We have implemented an efficient counting function for POGs and proved that
it computes that same quantity on the corresponding formulas:
\begin{lstlisting}
def count (pog : Pog) (nVars : Nat) (x : Var) : Nat := ...

theorem count_eq_countModels (pog : Pog) (nVars : Nat) (x : Var) :
  pog.count nVars x = (pog.toPropForm (.mkPos x)).countModels nVars
\end{lstlisting}
Here \lstinline{.mkPos x} denotes the positive literal for the variable \lstinline{x}. We have also proved that the calculation above really does
return the number of models:
\begin{lstlisting}
theorem countModels_eq_card_models {φ : PropForm ν} {s : Finset ν}
    (hvars : φ.vars ⊆ s) (hpar : φ.partitioned) :
  φ.countModels s.card = card (φ.models s)
\end{lstlisting}
In particular, taking \lstinline{s} to be exactly the variables of \lstinline{φ},
we have that the number of models on its variables is \lstinline{φ.countModels φ.vars.card}.

Given a weighting function \lstinline{weight : ν → Rat} that assigns a rational
number in $[0, 1]$ to each variable, we can calculate the weighted model count on
partitioned formulas as follows:
\begin{lstlisting}
def countWeight (weight : ν → Rat) : PropForm ν → Rat
  | tr         => 1
  | fls        => 0
  | var x      => weight x
  | neg φ      => 1 - φ.countWeight weight
  | disj φ ψ   => φ.countWeight weight + ψ.countWeight weight
  | conj φ ψ   => φ.countWeight weight * ψ.countWeight weight
  | impl _ _   => 0
  | biImpl _ _ => 0
\end{lstlisting}
We have implemented the corresponding count function on POGs and proved that
it calculates this quantity on the associated formulas,
though we have not yet proved the combinatorial lemma
that relates this calculation to the sum of the weights of the corresponding models.

\subsection{Verification specifics}
\label{subsection:verification:specifics}

In this subsection, we clarify what has been verified and what has been trusted.
Recall that our first step is to parse a CRAT file to read in the initial CNF formula
and the CRAT proof. We do not verify this step. There is no strong need to verify the
parsing of the proof, because if the checker claims success for \emph{any} proof, it does not really matter
whether the proof was the one the user intended.
On the other hand, it would be nice to have reassurance that the CNF formula that serves
as input to our verified checker matches the one described in the file.
An easy and fairly standard way of addressing this is to print out the CNF formula
right after parsing it and then do a diff with the original input.
This involves trusting only the correctness of the print procedure and diff.

Lean's code extraction replaces calculations with natural numbers and integers with
efficient but unverified arbitrary precision versions.
Lean also uses an efficient implementation of arrays; within the
formal system, these are defined in terms of lists, but code extraction replaces them
with dynamic arrays and uses reference counting to allow destructive updates when it is safe
to do so [cite counting immutable beans paper].
Code extraction also provides efficient implementations of tail recursive functions and loops
[cite ``do unchained'' paper].
Finally, we have implemented hashmaps in terms of arrays, but we have not yet proved
the correctness of all the primitive operations on these hashmaps.
Thus our proofs depend on the unverified assumption that those operations have the
expected properties.

In sum, in addition to trusting Lean's foundation and trusted kernel,
we also have to trust that code extraction respects that foundation,
that the implementation of arrays and hashmaps satisfy their internal descriptions,
and that, after parsing, the computation has the correct input formula.
All of our specifications have been completely proved and verified relative to that.

\subsection{Evaluation}

[TODO: Say something to establish that the procedure run in reasonable time and produce useful results; or maybe this does in the next section?]
